\pkg{Rgbp}: An R Package for Hierarchical Modeling and Method Checking

### 2013-05-29 jss/article.tex
biblio plus edit to intro and three data

%% almost as usual
\author{Joseph Kelly\\Harvard University \And 
             Carl Morris\\ Harvard University\And
             Hyungsuk Tak\\Harvard University }
\title{\pkg{Rgbp}: Bayesian Hierarchical Modeling and Frequentist Method Check}

@@

\section[introduction]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.
\pkg{Rgbp} uses Bayesian machinery to estimate a two-level model (a random-effects mixed model) and allows for a check of its frequentist properties via a repeated sampling procedure (which we call a ``method checking"). It is found that even in small samples our procedure yields good frequency properties. Also, this package will be useful for Bayesians who want to see a non-informative reference point before and after constructing their full-Bayesian hierarchical model. For frequentists, it will provide confidence intervals of a random-effect mixed model with good repeated sampling properties.

\section[Feasible Data Types]{Three Feasible Types of Data }
This package is intended to fit a multilevel model on the group-level (or unit-level) data in which each group-level observation is believed to have the Normal, Poisson, or Binomial distribution. In this section, we will introduce three specific types of feasible data sets.

\subsection{Normal: 8 Schools}
Education Testing Service conducted randomized experiments in eight separate schools and obtained this data set. It contains the coaching effects on SAT scores ($y_{j}, j=1, \ldots, 8$) and standard errors ($se_{j}, j=1, \ldots, 8$) of eight schools (Rubin, 1981).
\begin{CodeChunk}
\begin{CodeInput}
R> y  <- c(12, -3, 28,  7,  1,  8, 18, -1)
R> se <- c(18, 16, 15, 11, 11, 10, 10,  9)
\end{CodeInput}
\end{CodeChunk}

Each school's coaching effect has approximately Normal sampling distribution with known sampling variance, \emph{i.e.}, standard error of each school is assumed to be known or to be accurately estimated. So, it is reasonable to think that each school-level coaching effect is distributed as an independent Normal distribution given the unknown mean $\mu_{j}$ and known standard error:  $y_{j}\vert\mu_{j}\stackrel{ind}{\sim} \textrm{Normal}(\mu_{j}, se^{2}_{j}),~ j=1, \ldots, 8$. \pkg{Rgbp} includes this data set and can be called by typing `\code{R> data(schools)}' on \proglang{R}.

\subsection{Poisson: 31 Hospitals}
This data set is about the medical profiling evaluations for Coronary Artery Bypass Graft (CABG) surgeries of 31 New York hospitals conducted in 1992 (Morris and Lysy, 2012). It comprises of the number of deaths within a month of CABG surgeries in each hospital ($z_{j},~j=1, \ldots, 31$) and total number of patients receiving CABG surgeries (case load) in each hospital ($n_{j},~j=1, \ldots, 31$). The below code is an example of input based on the last ten hospital data.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c( 14,   9,  15,  13,  35,  26,  25,  20,   35,   27)
R> n <- c(593, 602, 629, 636, 729, 849, 914, 940, 1193, 1340)
\end{CodeInput}
\end{CodeChunk}

Considering the type of data, it makes sense to assume the number of deaths in each hospital has independent Poisson distribution given the unknown true rate parameter $\lambda_{j}$: $z_{j}\vert \lambda_{j}\stackrel{ind}{\sim} \textrm{Poisson}(n_{j}\lambda_{j})$, $j=1, \ldots, 31,$ where $n_{j}$ can be interpreted as an exposure (not necessarily an integer) and $\lambda_{j}$ as a true death rate per exposure. This data set is also included in the package and can be called by `\code{R> data(hospital)}' on \proglang{R}.

\subsection{Binomial: 18 Baseball Hitters}
This data set contains information about batting averages of 18 major league baseball players through their first 45 official at-bats of the 1970 season (Efron and Morris, 1975). Also, it has one covariate, Position, showing in which position each player was playing. For convenience, we transform this variable into a binary indicator, which is 1 if a player was a outfielder and 0 otherwise. The code below shows a way to make inputs. If we have more than one covariate, for example, \code{x1} and \code{x2}, then `\code{R> x <- cbind(x1, x2)}' will be the right input of the \code{gbp} function.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c(18, 17, 16, 15, 14, 14, 13, 12, 11, 11, 10, 10, 10, 10, 10,  9,  8,  7)
R> n <- c(45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45)
R> x <- c( 1,  1,  1,  1,  1,  0,  0,  0,  0,  1,  0,  0,  0,  1,  1,  0,  0,  0) 
\end{CodeInput}
\end{CodeChunk}
The data indicate that independent Binomial distribution is appropriate for each player's number of hits among 45 at-bats conditioning on the unknown true batting average $p_{j}$: $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(n_{j}, p_{j}), ~j=1, \ldots, 18$. This data set is also a part of the package and can be called on \proglang{R} by `\code{R> data(baseball)}'.

\section[Multilevel Structure]{Multilevel Structure}
Our multilevel model, also called a conditionally independent hierarchical model (Kass and Steffey, 1989), is a very powerful tool for exploring the hierarchical structure in data. For example, we can think about a district-level hierarchy (bigger population) for 8 schools, the state-level hierarchy for 31 hospitals, and the position-level hierarchy for 18 baseball players. \code{gbp}, one of functions in \pkg{Rgbp}, fits such a hierarchical model whose first-level hierarchy has a distribution of observed data and second-level (bigger population hierarchy) has a conjugate prior distribution on the first-level parameter. Users can determine one of three types of multilevel models, such as Normal-Normal, Poisson-Gamma, and Binomial-Beta, based on their data sets. 
\\

@@

Here the adjustment for density maximization (Morris and Tang, 2011), called ADM, comes. It assumes the Beta distribution for shrinkage because this distribution is also defined between 0 and 1. In the end, ADM  estimates its posterior moments, \emph{i.e.}, $E(B_{j}\vert\textrm{data})$ and $Var(B_{j}\vert\textrm{data})$, without any trouble that MLE can cause. Please refer to Morris and Tang (2011) for additional advantages of ADM.
\\

@@

Morris and Christiansen (1995) also looked into a similar ranking problem in hierarchical modeling, taking shrinkage into account.
\\

@@

Efron and Morris (1975) said ``no,'' and suggested James-Stein estimator, making each players' sample mean shrink towards an estimated mean of the second-level prior distribution. They showed that this estimator was more than three times efficient than the sample mean under squared error loss. On top of that, a few years later Morris (1983) introduced interval estimates that could successfully catch the batting averages over the remaining season, which MLE and its interval estimate could not.
\\

@@

\section[Reference]{Reference}
1. Christiansen, C. and Morris, C. (1996). ``Fitting and Checking a Two-Level Poisson Model: Modeling Patient Mortality Rates in Heart Transplant Patients,'' in \emph{Bayesian Biostatistics}, eds. D. Berry and D. Stangl, New York: Marcel Dekker, pp. 467-561.
\\

2. Efron, B. and Morris, C. (1975). ``Data Analysis Using Stein's Estimator and its Generalizations.'' \emph{Journal of the American Statistical Association}. \textbf{70}. 311-319.
\\

3. Kass, R. and Steffey, D. (1989). ``Approximate Bayesian Inference in Conditionally Independent Hierarchical Models (Parametric
Empirical Bayes Models).'' \emph{Journal of the American Statistical Association}. \textbf{84}. 717-726.
\\

4. Morris, C. (1983). ``Parametric Empirical Bayes Inference: Theory and Applications.'' \emph{Journal of the American Statistical Association}. \textbf{78}. 47-55.
\\

5. Morris, C. and Christiansen, C. (1995). ``Hierarchical Models for Ranking and for Identifying Extremes, With Application,'' in \emph{Bayesian Statistics 5}, eds. J. Bernardo, J. Berger, A. Dawid, and A. Smith, New York: Oxford University Press, pp. 277-296.
\\

6. Morris, C. and Tang, R. (2011). ``Estimating Random Effects via Adjustment for Density Maximization.'' \emph{Statistical Science}. \textbf{26}. 271-287.
\\

7. Morris, C. and Lysy, M. (2012). ``Shrinkage Estimation in Multilevel Normal Models.'' \emph{Statistical Science}. \textbf{27}. 115-134.
\\

8. Rubin, D. B. (1981). ``Estimation in parallel randomized
  experiments.'' \emph{Journal of Educational Statistics}. \textbf{6}. 377-401.
\\

>>>>

%% almost as usual
\author{Joseph Kelly\\Harvard University \And 
             Carl Morris\\ Harvard University\And
             Hyungsuk Tak\\Harvard University }
\title{\pkg{Rgbp}: An R Package for Hierarchical Modeling and Method Checking}

@@

\section[introduction]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.
\pkg{Rgbp} is an R package for estimating and validating a two-level model (a random-effects mixed model). The estimation procedure utilizes Bayesian machinery and the validation involves checking frequency properties of the procedure via repeated sampling (which we call ``method checking"). It is found that even in small samples our procedure yields good frequency properties in comparison to other methods such as Maximum Likelihood Estimation (MLE). This package will be useful for frequentists and Bayesians alike. Bayesians are able to use the package to see a non-informative reference point before and after constructing their full-Bayesian hierarchical model and frequentists, now have a procedure that will provide confidence intervals of a random-effect mixed model with good repeated sampling properties.

%TODO: Do we want this here? Or just go straight into theory then follow with examples?
\section[Feasible Data Types]{Three Feasible Types of Data }
\pkg{Rgbp} is intended to fit a model where two levels of structure are assumed (unit and group levels). The model can be characterized by the distributional family assumed for the group level data and in the case of \pkg{Rgbp} this will be either the Normal, Poisson, or Binomial distribution. In this section, we will introduce three specific types of feasible data sets that illustrate this fact.

\subsection{Normal: 8 Schools}
The Education Testing Service (ETS) conducted randomized experiments in eight separate schools (group) to test whether students (unit) SAT scores are effected by coaching. The dataset contains the estimated coaching effects on SAT scores ($y_{j}, j=1, \ldots, 8$) and standard errors ($se_{j}, j=1, \ldots, 8$) of the eight schools \cite{1981}.
\begin{CodeChunk}
\begin{CodeInput}
R> y  <- c(12, -3, 28,  7,  1,  8, 18, -1)
R> se <- c(18, 16, 15, 11, 11, 10, 10,  9)
\end{CodeInput}
\end{CodeChunk}

Due to the nature of the test each school's coaching effect has an approximately Normal sampling distribution with known sampling variance, \emph{i.e.}, standard error of each school is assumed to be known or to be accurately estimated. So, it is reasonable to think that each school-level coaching effect is distributed as an independent Normal distribution given the unknown mean $\mu_{j}$ and known standard error:  $y_{j}\vert\mu_{j}\stackrel{ind}{\sim} \textrm{Normal}(\mu_{j}, se^{2}_{j}),~ j=1, \ldots, 8$. \pkg{Rgbp} includes this data set and can be called by the command `\code{data(schools)}' on \proglang{R}.

\subsection{Poisson: 31 Hospitals}
The following data are from medical profiling evaluations for Coronary Artery Bypass Graft (CABG) surgeries of 31 New York hospitals conducted in 1992 \cite{2012}. It comprises of the number of deaths within a month of CABG surgeries in each hospital ($z_{j},~j=1, \ldots, 31$) and the total number of patients receiving CABG surgeries (case load) in each hospital ($n_{j},~j=1, \ldots, 31$). The following code is an example of input based on the last ten hospital data.
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c( 14,   9,  15,  13,  35,  26,  25,  20,   35,   27)
R> n <- c(593, 602, 629, 636, 729, 849, 914, 940, 1193, 1340)
\end{CodeInput}
\end{CodeChunk}
Considering the type of data, it is reasonable to assume the number of deaths in each hospital follow independent Poisson distributions given an unknown true rate parameter $\lambda_{j}$: $z_{j}\vert \lambda_{j}\stackrel{ind}{\sim} \textrm{Poisson}(n_{j}\lambda_{j})$, $j=1, \ldots, 31,$ where, for each $j$, $n_{j}$ can be interpreted as an exposure (not necessarily an integer) and $\lambda_{j}$ as the true death rate per exposure. This data set is also included in the package and can be called by `\code{data(hospital)}' on \proglang{R}.

\subsection{Binomial: 18 Baseball Hitters}
The following dataset contains information on the batting averages of 18 major league baseball players through their first 45 official at-bats of the 1970 season \cite{1975}. In addition one covariate relating to the position each player was playing was observed and for illustrative purposes, we transform this variable into a binary indicator (1 if a player was a outfielder and 0 otherwise).
\begin{CodeChunk}
\begin{CodeInput}
R> z <- c(18, 17, 16, 15, 14, 14, 13, 12, 11, 11, 10, 10, 10, 10, 10,  9,  8,  7)
R> n <- c(45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45, 45)
R> x <- c( 1,  1,  1,  1,  1,  0,  0,  0,  0,  1,  0,  0,  0,  1,  1,  0,  0,  0) 
\end{CodeInput}
\end{CodeChunk}
The data indicate that independent Binomial distribution is appropriate for each player's number of hits among 45 at-bats conditioning on the unknown true batting average $p_{j}$: $z_{j}\vert p_{j}\stackrel{ind}{\sim} \textrm{Binomial}(n_{j}, p_{j}), ~j=1, \ldots, 18$. This data set is also a part of the package and can be called on \proglang{R} by `\code{data(baseball)}'.

\section[Multilevel Structure]{Multilevel Structure}
Our multilevel model, also called a conditionally independent hierarchical model \cite{1989}, is a very powerful tool for exploring the hierarchical structure in data. For example, we can think about a district-level hierarchy (bigger population) for 8 schools, the state-level hierarchy for 31 hospitals, and the position-level hierarchy for 18 baseball players. \code{gbp}, one of functions in \pkg{Rgbp}, fits such a hierarchical model whose first-level hierarchy has a distribution of observed data and second-level (bigger population hierarchy) has a conjugate prior distribution on the first-level parameter. Users can determine one of three types of multilevel models, such as Normal-Normal, Poisson-Gamma, and Binomial-Beta, based on their data sets. 
\\

@@

Here the adjustment for density maximization \cite{2011}, called ADM, comes. It assumes the Beta distribution for shrinkage because this distribution is also defined between 0 and 1. In the end, ADM  estimates its posterior moments, \emph{i.e.}, $E(B_{j}\vert\textrm{data})$ and $Var(B_{j}\vert\textrm{data})$, without any trouble that MLE can cause. Please refer to \cite{2011} for additional advantages of ADM.
\\

@@

\cite{1995} also looked into a similar ranking problem in hierarchical modeling, taking shrinkage into account.
\\

@@

\cite{1975} said ``no,'' and suggested James-Stein estimator, making each players' sample mean shrink towards an estimated mean of the second-level prior distribution. They showed that this estimator was more than three times efficient than the sample mean under squared error loss. On top of that, a few years later \cite{1983} introduced interval estimates that could successfully catch the batting averages over the remaining season, which MLE and its interval estimate could not.
\\

@@

\bibliography{bibliography}
